Composable data visualizations

### 2020-06-12 wip-compost/paper.tex
Final edits

\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}

@@

\section{Introduction}
Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
a bar chart that shows two different values for each bar. The chart on the right consists of two
line charts that share the X axis with parts of the timeline highlighted using two different colors.

There is a plenty of libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario.
Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
way for adding a background, or sharing an axis between charts. The alternative is to use a more
low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use ggplot2 \cite{ggplot2}, based on the Grammar of Graphics \cite{grammar}.
A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
properties (X and Y coordinate, shape, color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

What would an elegant functional approach to data visualization look like? A functional programmer
would want a domain-specific language that has a small number of primitives; allow us to define
high level abstractions such as a bar chart and its basic building blocks are expressed in terms
of domain values such as the exchange rate, rather than pixels.

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about the UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

As is often the case with domain-specific languages, finding the right primitives is more of an art
than science. For this reason, we present an answer, a library named Compost, as a functional pearl.
We hope to convince the reader that Compost is elegant and we illustrate this with a wide range
of examples. Compost has a number of specific desirable properties:

\begin{itemize}
\item Charts are composed from a small number of primitive building blocks using a small number of
  combinators. In particular, concepts such as bar charts, line charts or charts with aligned
  axes are all expressed in terms of more basic concepts.
\item The primitives are specified in domain terms. When drawing a line, the value of an Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
\item Most common chart types can be easily captured as high level abstractions, but there is an
  elengant way of creating a majority of more interesting custom charts.
\item The approach can easily be integrated with the Elm architecture \cite{elm}
	to creating web-based charts that involve animations or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. We outline the structure of an implementation, but omit the details. Filling
those in requires careful thinking about geometry and projections, but there are no unexpected
surprises. A complete F\# implementation, including the examples used in this paper, is available
at: \urrl{http://github.com/compostjs}.

@@

\begin{figure}[t]
\includegraphics[scale=1,trim={0cm 6.5cm 6cm 0cm},clip]{figures/values.pdf}
\caption{On a continuous scale (above), an exact position is determined by a number.
  On a categorical scale (below), an exact position is determined by the category and a
  numerical ratio from 0 to 1.}
\label{fig:scales}
\end{figure}

@@

In the election results chart in Figure~\ref{fig:charts} (left), the X axis shows categorical
values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
Y axis shows numerical values representing the number of seats won such as $\num{365}$ MPs.
When creating data visualizations, those are the values that the user needs to specify. This is
akin to most high-level charting libraries such as Google Charts, but in contrast with more
flexible libraries like D3.

Our design focuses on two-dimensional charts with X and Y axes. Values mapped to those axes
can be either categorical (e.g.~political parties, countries) or continuous
(e.g.~number of votes, exchange rates). The mapping from categorical and continuous values
to positions on the chart is done automatically. For continuous values, this involves
applying a linear transformation. For categorical values, the mapping is more difficult.

For example, in the UK election results chart, the X axis is categorical. The library automatically
divides the available space between the six categorical values (political parties). The value
\strf{Green} does not determine an exact position on the axis, but rather a range. To determine
an exact position, we also need to attach a value between $\num{0}$ and $\num{1}$ to the
categorical value. This identifies a relative position in the available range.

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$, consisting
of a categorical value $c$ (implemented as a string) and a ratio $r$ between $0$ and $1$.
%
\begin{figure}
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \\
  & | & \kvd{cont}~n\\
  ~
\end{array}
\qquad
\begin{array}{rclcl}
s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
 & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
 & | & \kvd{bubble}~\gamma, v_x, v_y, w, h
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

@@

Now that we know how Compost represents values, we can define the basic elements of its
domain-specific langauge. A chart is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by X and Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels.

Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
which overlays all shapes from a given list. When doing this, Compost automatically infers the
scales of X and Y axes and calculates suitable projections using a method discussed in the next
section. Finally, $\kvd{padding}$ adds padding around a specified shape and $\kvd{axis}$ adds
an axis showing the inferred scale on the left, right, top or bottom of a given shape.
Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left).
We use the \fkvd{let} construct of the host functional language to structure the code:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{conservative},~\ident{labour}~=\\
\quad \kvd{fill}~\strf{\#0000ff},
 ~[\;\,(\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{365}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{365}), (\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{0})\;\,],\\
\quad \kvd{fill}~\strf{\#ff0000},
 ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]\\[0.5em]
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{labour}, \ident{conservative}~]))\\
\end{array}
\end{equation*}

\noindent
The chart specification overlays two bars of different colors and then adds axes to the bottom and
left of the chart. The two bars are filled rectangles defined using four corner points. The Y
coordinates are specified as continuous values, while the X coordinates are categorical. For
the Conservative party, two of the points have the Y coordinate set to $\kvd{cont}~\num{0}$ (bottom of the bar)
and two have the Y coordinate set to $\kvd{cont}~\num{365}$ (top of the bar). The two X coordinates
are the start and the end of the range allocated for the \strf{Conservative} category,
i.e.~$\kvd{cat}~\strf{Conservative}, \num{0}$ on the left and $\kvd{cat}~\strf{Conservative}, \num{1}$
on the right.

@@

Extending the snippet to generate a grouped bar chart that shows two results
for each party as in Figure~\ref{fig:charts} is easy. Given a party $p$, we need to
generate two rectangles, one with X coordinates $\kvd{cat}~p, 0$ and $\kvd{cat}~p, 0.5$
and the other with X coordinates $\kvd{cat}~p, 0.5$ and $\kvd{cat}~p, 1$.
In the following snippet, we use a \fkvd{for} comprehension to generate the list. All remaining
constructs are primitives of the Compost domain-specific language. Assuming \ident{elections} is a
list of election results containing a five-element tuple consisting of a party name, colors for
2017 and 2019 and results for 2017 and 2019 we create the chart using:
%
\begin{equation*}
\begin{array}{l}
  \kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
  \quad\quad\quad \kvd{fill}~\ident{clr17},~[\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,], \\
  \quad\quad\quad \kvd{fill}~\ident{clr19},~[\,(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})\,]~~]~~]\,)\,)\\
\end{array}
\end{equation*}

\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars. A padding is specified in pixels rather than in terms of domain values. This is
sometimes preferrable over, for example, drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$.
The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.

@@

When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
to position the child elements relatively to each other. The Compost library positions the elements
automatically. This is done in two steps. First, Compost infers the \emph{scales} for X and Y
axes. A scale represents the range of values that needs to fit in the space available for the chart.
Second, Compost calculates a \emph{projection}, a mapping from domain-specific values of the scale
to the available screen space. A scale $l$ is defined in Figure~\ref{fig:scale}.

@@

A continuous scale is defined by a minimal and maximal value that need to be mapped to the
available chart space. A categorical scale is defined by a list of individual categorical values.
Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
will use an equal space for each category, regardless of where in this space a shape needs to
appear.

Inferring scales is done by a simple recursive function that walks over the given shape and
constructs two scales for the X and Y axis, using the X and Y coordinates that appear in the shape.
Most of the work is done by a simple helper function that takes two scales, $l_1$ and $l_2$,
and produces a new scale that represents the union of the two:
%
\begin{equation*}
\begin{array}{l}
\ident{union}~(\kvd{continuous}~n_l, n_h)~(\kvd{continuous}~n'_l, n'_h) =\\
\qquad\kvd{continuous}~\min(n_l, n'_l), \max(n_h, n'_h)\\[0.5em]
\ident{union}~(\kvd{categorical}~[\,c_1, \ldots, c_p\,])~(\kvd{categorical}~[\,c'_1, \ldots, c'_q\,]) =\\
\qquad \kvd{categorical}~[\,c_1, \ldots, c_p\,] \;@\; [\,c'_i~|~ \forall i\in 1\ldots q, \nexists j. c_j = c'_i \;]
\end{array}
\end{equation*}

@@

Once Compost computes scales for a given shape and its sub-shapes, it constructs a projection
that maps domain-specific values to the available chart space. We discuss this in
Section~\ref{sec:impl}. For a continuous scale, the projection is a linear transformation. For
categorical scale with $k$ values, we split the available chart space into $k$ equally sized
regions and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
according to the ratio $r$.

\subsection{Types and units of measure}
We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
which types could make composing charts in Compost safer. First, a type representing a shape could
specify whether the X and Y axes represent categorical or continous values. This would rule out
mixing of different values on a single scale and guarantee that the \ident{union} operation,
sketched in the previous section, is never called in a way leading to an undefined result.
Second, the type of values mapped to an axis could be further annotated with units of measure
\cite{units}. Using the F\# notation where $n\langl u\rangl$ is a number $n$ with unit $u$,
an axis containing a value $\kvd{cont}~\num{317}\langl \ident{mp}\rangl$ would then be incompatible
with an axis containing a value $\kvd{cont}~\num{1.32}\langl \ident{gbp}/\ident{usd}\rangl$.

The design of the type system for Compost is straightforward, so we only present a brief sketch.
There are two kinds of types; $\sigma$ is a type of values and $\tau$ is a type of shapes.
Assuming $u$ denotes a unit of measure, the types are defined as:
%
\begin{equation*}
\begin{array}{rcl}
\sigma & = & \kvd{Cat}~u \lsep \kvd{Cont}~u
\end{array}
\qquad\qquad
\begin{array}{rcl}
\tau & = & \kvd{Shape}~\sigma_x, \sigma_y \\
\end{array}
\end{equation*}

\vspace{-1em}
\noindent
Correspondingly, there are two kinds of judgements; $v\vdash \sigma$ indicates the type of a
value, while $s\vdash \tau$ indicates the type of a shape. The typing rules for two of the
basic chart primitives, \kvd{line} and \kvd{overlay} look as follows:
%
\begin{equation*}
\inference
  {v_{xi} \vdash \sigma_x & v_{yi} \vdash \sigma_y}
  {\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
\qquad
\inference
  {s_i \vdash \kvd{Shape}~\sigma_x, \sigma_y}
  {\kvd{overlay}~[\,s_1, \ldots, s_n\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
\end{equation*}

\vspace{-0.5em}
\noindent
The rule for \kvd{line} ensures that all X and Y values have the same types, $\sigma_x$ and $\sigma_y$,
respectively and infers $\kvd{Shape}~\sigma_x, \sigma_y$ as the type of the shape. The rule for
\kvd{overlay} ensures that all composed shapes have the same type, including the type of X and Y scales.

@@

Most charts have one X and one Y axis that are determined by the values the chart shows,
but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
has two different Y axes, one for GBP/USD and one for GBP/EUR. In the next two sections, we look at
three combinators that control the scale inferrence process and what flexibility this enables.

@@

>>>>

\newcommand{\X}{\emph{x}\ }
\newcommand{\Y}{\emph{y}\ }
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\fkvd}[1]{\textnormal{\textcolor{fkvdclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}
\newcommand{\urrl}[1]{\textnormal{\textcolor{urlclr}{\url{#1}}}}

@@

\section{Introduction}
Let's say we want to create the two charts in Figure~\ref{fig:charts}. The chart on the left is
a bar chart that shows two different values for each bar. The chart on the right consists of two
line charts that share the \X axis with parts of the timeline highlighted using two different colors.

Many libraries that can draw bar charts and line charts, but adding those extra
features will only be possible if the author already thought about your exact scenario.
Google Charts \cite{gcharts} supports the left chart (it is called Dual-X Bar Chart) but there is no
way to add a background, or sharing an axis between charts. The alternative is to use a more
low-level library. In D3 \cite{d3} you construct the chart piece by piece, but you have to
tediously transform your values to coordinates in pixels yourself. For scientific plots,
you could use ggplot2 \cite{ggplot2}, based on the Grammar of Graphics \cite{grammar}.
A chart is a mapping from data to geometric objects (points, bars, lines) and their visual
properties (\X and \Y coordinate, shape, color). However, the range of charts that can be
created using this systematic approach is still somewhat limited.

What would an elegant functional approach to data visualization look like? A functional programmer
would want a domain-specific language that has a small number of primitives that allow us to define
high level abstractions such as a bar chart and that uses domain values such as the exchange
rate, rather than pixels, in its basic building blocks.

\begin{figure}[h]
  \includegraphics[scale=0.57]{figures/charts}
  \vspace{0.25em}
  \caption{Two charts about UK politics: Comparison of election results from 2017 and 2019 (left)
    and GBP/USD exchange rate with highlighted areas before and after the 23 June 2016 Brexit vote.}
  \label{fig:charts}
\end{figure}

As is often the case with domain-specific languages, finding the right primitives is more of an art
than science. For this reason, we present our solution, a library named Compost, as a functional pearl.
We hope to convince the reader that Compost is elegant and we illustrate this with a wide range
of examples. Compost has a number of specific desirable properties:

\begin{itemize}
\item Concepts such as bar charts, line charts or charts with aligned
  axes are all expressed in terms of more primitive building blocks using a small number of combinators.
\item The primitives are specified in domain terms. When drawing a line, the value of a \\ \Y coordinate
  is an exchange rate of 1.36 USD/GBP, not 67 pixels from the bottom.
\item Most common chart types can be easily captured as high level abstractions, but there is a
  way of creating a majority of more interesting custom charts.
\item The approach can easily be integrated with the Elm architecture \cite{elm}
	to creating web-based charts that involve animations or interaction with the user.
\end{itemize}
%
The presentation in this paper focuses on explaining the primitives and combinators of the
domain-specific language. We outline the structure of an implementation, but omit the details; filling
those in merely requires careful thinking about geometry and projections. An F\#
implementation, including all our examples, is available
at: \urrl{http://github.com/compostjs}.

@@

\begin{figure}[t]
\includegraphics[scale=1,trim={0cm 6.9cm 6cm 0cm},clip]{figures/values.pdf}
\caption{On a continuous scale (above), an exact position is determined by a number.
  On a categorical scale (below), an exact position is determined by the category and a
  number between 0 and 1.}
\label{fig:scales}
\end{figure}

@@

In the election results chart in Figure~\ref{fig:charts} (left), the \X axis shows categorical
values representing the political parties such as \strf{Conservative} or \strf{Labour}. The
\Y axis shows numerical values representing the number of seats won such as $\num{365}$ MPs.
When creating data visualizations, those are the values that the user needs to specify. This is
akin to most high-level charting libraries such as Google Charts, but in contrast with more
flexible libraries like D3.

Our design focuses on two-dimensional charts with \X and \Y axes. Values mapped to those axes
can be either categorical (e.g.~political parties, countries) or continuous
(e.g.~number of votes, exchange rates). The mapping from categorical and continuous values
to positions on the chart is done automatically. We discuss this in Section~\ref{sec:basic-scales}.

For example, in the UK election results chart, the \X axis is categorical. The library automatically
divides the available space between the six categorical values (political parties). The value
\strf{Green} does not determine an exact position on the axis, but rather a range. To determine
an exact position, we also need to attach a value between $\num{0}$ and $\num{1}$ to the
categorical value. This identifies a relative position in the available range.

Figure~\ref{fig:scales} illustrates the two kinds of values using the axes from the UK
election results chart. In Figure~\ref{fig:shape}, we define a value $v$ as either a continuous value
$\kvd{cont}~n$ containing any number $n$ or a categorical value $\kvd{cat}~c, r$ consisting
of a categorical value $c$ and a number $r$ between $0$ and $1$. As discussed in
Section~\ref{sec:basic-types}, continuous and categorical values can also be annotated with
units of measure to make the values more descriptive.
%
\begin{figure}
\begin{equation*}
\begin{array}{rcl}
v & = & \kvd{cat}~c, r \\
  & | & \kvd{cont}~n\\[1.25em]
  ~
\end{array}
\qquad
\begin{array}{rclcl}
s & = &\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | & \kvd{overlay}~[\,s_1, \ldots, s_n\,]\\
 & | & \kvd{fill}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] & | &\kvd{axis}_{l/r/t/b}~s\\
 & | & \kvd{text}~\gamma, v_x, v_y, t & | &\kvd{padding}~n_t,n_r,n_b,n_l,s\\
 & | & \kvd{bubble}~\gamma, v_x, v_y, w, h
\end{array}
\end{equation*}
\caption{Core primitives of the Compost domain-specific language. Values $v$ are either categorical
  or continuous; a shape $s$ is then defined as a simple recursive algebraic data type.}
\label{fig:shape}
\end{figure}

@@

Now we can define the besic elements of the Compost domain-specific langauge.
A chart element is represeted by the shape $s$ defined in Figure~\ref{fig:shape}.
A primitive shape can be a text label, a line connecting a list of points, a filled polygon
defined by a list of points or a bubble at a given point with a given width and height.
The position of points is specified by \X and \Y coordinates, which can be either categorical or
continuous values. For text, line, polygon and bubble, we also include a parameter $\gamma$
that specifies the element color. The width and height of a bubble is given in pixels rather than
in domain units.

Figure~\ref{fig:shape} also defines three combinators. The most important is $\kvd{overlay}$,
which overlays given shapes. When doing this, Compost infers the range of values on the \X and \Y
axes and calculates suitable projections using a method discussed in the next
section. The $\kvd{padding}$ combinator adds padding around a specified shape and $\kvd{axis}$ adds
an axis showing the inferred scale on the left, right, top or bottom of a given shape.
Using those primitives, we can construct the simple UK election results bar chart in Figure~\ref{fig:simple} (left):
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{conservative},~\ident{labour}~=\\
\quad \kvd{fill}~\strf{\#0000ff},
 ~[\;\,(\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Conservative}, \num{0}), (\kvd{cont}~\num{365}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{365}), (\kvd{cat}~\strf{Conservative}, \num{1}), (\kvd{cont}~\num{0})\;\,],\\
\quad \kvd{fill}~\strf{\#ff0000},
 ~[\;\,(\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\strf{Labour}, \num{0}), (\kvd{cont}~\num{202}),\\
\hspace{6.7em}\;(\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{202}), (\kvd{cat}~\strf{Labour}, \num{1}), (\kvd{cont}~\num{0})\,\;]\\[0.5em]
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\ident{labour}, \ident{conservative}~]))\\
\end{array}
\end{equation*}

\vspace{-0.25em}
\noindent
We use the \fkvd{let} construct of the host functional language to structure the code. The chart
specification overlays two bars of different colors and then adds axes to the bottom and left of the chart.
The two bars are filled rectangles defined using four corner points. The \Y
coordinates are specified as continuous values, while the \X coordinates are categorical. For
the Conservative party, two of the points have the \Y coordinate set to $\kvd{cont}~\num{0}$ (bottom of the bar)
and two have the \Y coordinate set to $\kvd{cont}~\num{365}$ (top of the bar). The two \X coordinates
are the start and the end of the range allocated for the \strf{Conservative} category,
i.e.~$\kvd{cat}~\strf{Conservative}, \num{0}$ on the left and $\kvd{cat}~\strf{Conservative}, \num{1}$
on the right.

@@

Extending the snippet to generate a grouped bar chart that shows two results
for each party as in Figure~\ref{fig:charts} is not much harder. Given a party $p$, we need to
generate two rectangles, one with \X coordinates $\kvd{cat}~p, 0$ and $\kvd{cat}~p, 0.5$
and the other with \X coordinates $\kvd{cat}~p, 0.5$ and $\kvd{cat}~p, 1$.
In the following snippet, we use a \fkvd{for} comprehension to generate the list. All remaining
constructs are primitives of the Compost domain-specific language. Assuming \ident{elections} is a
list of election results containing a five-element tuple consisting of a party name, colors for
2017 and 2019 and results for 2017 and 2019 we create the chart using:
%
\begin{equation*}
\begin{array}{l}
  \kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[\\
  \quad \fkvd{for}~\ident{party}, \ident{clr17}, \ident{clr19}, \ident{mp17}, \ident{mp19}~\fkvd{in}~\ident{elections}~\rightarrow\\
  \quad\quad \kvd{padding}~0,10,0,10,\,\kvd{overlay}~[\\
  \quad\quad\quad \kvd{fill}~\ident{clr17},~[\,(\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0}), (\kvd{cont}~\ident{mp17}),\\
  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp17}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}) \,], \\
  \quad\quad\quad \kvd{fill}~\ident{clr19},~[\,(\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\num{0}), (\kvd{cat}~\ident{party}, \num{0.5}), (\kvd{cont}~\ident{mp19}),\\
  \quad\quad\quad \hspace{4.08em}           ~(\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\ident{mp19}), (\kvd{cat}~\ident{party}, \num{1}), (\kvd{cont}~\num{0})\,]~~]~~]\,)\,)\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
Aside from iterating over all available parties and splitting the bar, the example also adds padding
around the bars. This is specified in pixels rather than in terms of domain values. A similar
result could be be achieved by drawing a bar using a range from $\num{0.05}$ to $\num{0.5}$, but
specifying padding precisely in pixels is sometimes preferable.
The chart is still missing a title, which we add in Section~\ref{sec:abstractions}.

@@

We follow the terminology of Vega \cite{vega} and use the term \emph{scale} to refer to the mapping of
values to positions on a screen; a \emph{coordinate} is a value representing a position on a
scale and the term \emph{axis} is used to refer to the visual representation of a scale.

Scales are an important concept in Compost.
When composing shapes using the \kvd{overlay} primitive, the user does not need to specify how
to position the child elements relatively to each other. The Compost library positions the elements
automatically. This is done in two steps. During pre-processing, Compost infers the scales for
\X and \Y axes. A scale represents the range of values that needs to fit in the space available for
the chart. When rendering a shape, Compost projects domain-specific values to the available screen
space based on the inferred scale. A scale $l$ is defined in Figure~\ref{fig:scale}.
%
A continuous scale is defined by a minimal and maximal value that need to be mapped to the
available chart space. A categorical scale is defined by a list of individual categorical values.
Note that we do not need a minimal and maximal ratios of the used categorical values as Compost
will use an equal space for each category, regardless of where in this space a shape needs to
appear.

@@

Scale inference is done by a simple recursive function that walks over the given shape and
constructs two scales for the \X and \Y axis, using the \X and \Y coordinates that appear in the shape.
Most of the work is done by a simple helper function that takes two scales, $l_1$~and $l_2$,
and produces a new scale that represents the union of the two:
%
\begin{equation*}
\begin{array}{l}
\ident{union}~(\kvd{continuous}~n_l, n_h)~(\kvd{continuous}~n'_l, n'_h) =\\
\qquad\kvd{continuous}~\min(n_l, n'_l), \max(n_h, n'_h)\\[0.5em]
\ident{union}~(\kvd{categorical}~[\,c_1, \ldots, c_p\,])~(\kvd{categorical}~[\,c'_1, \ldots, c'_q\,]) =\\
\qquad \kvd{categorical}~[\,c_1, \ldots, c_p\,] \;@\; [\,c'_i~|~ \forall i\in 1\ldots q, \nexists j. c_j = c'_i \;]
\end{array}
\end{equation*}

@@

The scales inferred during pre-processing are later used when rendering a shape.
We discuss the implementation in Section~\ref{sec:impl}. The key operation is projection
which, given a coordinate, a scale and an area on the screen, produces a position on
the screen. For a continuous scale, the projection is a linear transformation. For
categorical scale with $k$ values, we split the available chart space into $k$ equally sized
regions and then map a categorical value $\kvd{cat}~c, r$ to the region corresponding to $c$
according to the ratio $r$.

\subsection{Types and units of measure}
\label{sec:basic-types}

We introduce the Compost domain-specific language as untyped, but there are some obvious ways in
which types can make composing charts in Compost safer. First, a type representing a shape could
specify whether the \X and \Y axes represent categorical or continous values. This would rule out
mixing of different values on a single scale and guarantee that the \ident{union} operation,
sketched in the previous section, is never called in a way leading to an undefined result.
Second, the type of values mapped to an axis could be further annotated with units of measure
\cite{units}. Using the F\# notation where $n\langl u\rangl$ is a number $n$ with unit $u$,
an axis containing a value $\kvd{cont}~\num{317}\langl \ident{mp}\rangl$ would then be incompatible
with an axis containing a value $\kvd{cont}~\num{1.32}\langl \ident{gbp}/\ident{usd}\rangl$.

We only outline the type system here. There are two kinds of types; $\sigma$ is a type of values
and $\tau$ is a type of shapes. Assuming $u$ denotes a unit of measure, the types are defined as:
%
\begin{equation*}
\begin{array}{rcl}
\sigma & = & \kvd{Cat}~u \lsep \kvd{Cont}~u
\end{array}
\qquad\qquad
\begin{array}{rcl}
\tau & = & \kvd{Shape}~\sigma_x, \sigma_y \\
\end{array}
\end{equation*}

\vspace{-1.25em}
\noindent
Correspondingly, there are two kinds of typing judgements; $v\vdash \sigma$ indicates the type of a
value, while $s\vdash \tau$ indicates the type of a shape. The typing rules for two of the
basic chart primitives, \kvd{line} and \kvd{overlay} look as follows:
%
\begin{equation*}
\inference
  {v_{xi} \vdash \sigma_x & v_{yi} \vdash \sigma_y}
  {\kvd{line}~\gamma, [\,v_{x1}, v_{y1}, \ldots, v_{xn}, v_{yn}\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
\qquad
\inference
  {s_i \vdash \kvd{Shape}~\sigma_x, \sigma_y}
  {\kvd{overlay}~[\,s_1, \ldots, s_n\,] \vdash \kvd{Shape}~\sigma_x, \sigma_y}
\end{equation*}

\vspace{-0.5em}
\noindent
The rule for \kvd{line} ensures that all X and Y values have the same types, $\sigma_x$ and $\sigma_y$,
respectively and infers $\kvd{Shape}~\sigma_x, \sigma_y$ as the type of the shape. The rule for
\kvd{overlay} ensures that all composed shapes have the same type, including the type of \X and \Y scales.

@@

Most charts have one \X and one \Y scale that are determined by the values the chart shows,
but there are interesting exceptions. The chart in Figure~\ref{fig:charts} (right)
has two different \Y axes, one for GBP/USD and one for GBP/EUR. In the next two sections, we look at
three combinators that control the scale inference process and what flexibility this enables.

@@

The most interesting primitive for controlling scale composition
is $\kvd{nest}_{x/y}$. The behaviour is similar to that of the SVG viewport \cite{svg}.
It takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments and
nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$.
When inferring scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
constructed from the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP/EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.
